Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-05-06 2015-eamt-melsa/poster/2015-eamt-melsa-poster.tex
stuff
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60275 72bbbca6-d526-0410-a7d9-f06f51895060

\settitle{ 
  \vbox{
    \color{titlefgcolor} {\bfseries \Huge  \@title \par}
    \vspace*{-0.5em}
    \begin{tabular}{llr}
    {\Large \@author \par} & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  &  \multirow{3}{*}{\@titlegraphic } \\ % [\TP@titlegraphictotitledistance]}  \\
    & ~ & ~   \\
    {\large \@institute}   & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ &  \\
    \end{tabular}
    \vspace*{1em}
  } 
}

@@

\title{Unsupervised training of maximum-entropy models 
 for lexical selection in rule-based machine translation}
 \institute{$^\star$UiT Norgga árktalaš universitehta, $^\dagger$Universitat d'Alacant} % See Section 4.1
\author{Francis M. Tyers,$^\star$ Felipe Sánchez-Martínez,$^\dagger$ Mikel L. Forcada$^\dagger$} 
\titlegraphic{
\includegraphics[width=0.04\textwidth]{LogoSamisk.png} \hspace{25pt} \includegraphics[width=0.15\textwidth]{logodlsioficialcolor.eps} \hspace{10pt} \includegraphics[width=0.05\textwidth,angle=90]{apertium5a.pdf}
}
%\usetheme{Basic} % See Section 5
%\usebackgroundstyle{Rays}
\begin{document}

@@

\block[titleinnersep=10em]{Lexical selection}{

@@

  \block[titleinnersep=10em]{Apertium (\url{www.apertium.org})}{
    \begin{center}
      \includegraphics[width=0.45\textwidth]{architecture-lexsel.pdf}
    \end{center}
    \begin{itemize}
      \item Free/open-source rule-based MT platform
      \item Around 40 released language pairs
    \end{itemize}
  }
% -- weights should go on operations
%  <rule weight="0.906306">
%    <match lemma="ondare" tags="n"><select lemma="acervo" tags="n.*"/></match>
%    <match lemma="historiko" tags="adj.*"/>
%  </rule>
%  <rule weight="1.08987">
%    <match lemma="ondare" tags="n"><select lemma="patrimonio" tags="n.*"/></match>
%    <match lemma="historiko" tags="adj.*"/>
%  </rule>
%

  \block[titleinnersep=10em]{Training schema}{

%\begin{displaymath}
%    \label{eq:steps}
%    S \to \mbox{\framebox{\parbox{4.0cm}{\textit{pre-lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{3.5cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{4.0cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
%  \end{displaymath}

  \begin{center}
    \includegraphics[width=0.40\textwidth]{schema.eps}
  \end{center}
  }

@@

\block[titleinnersep=10em]{Training}{ % See Section 4.2

Translate SL sentences to TL using existing MT system and score the translations using an $n$-gram TL model.

\useinnerblockstyle{Table}
\begin{center}
\begin{tabular}{|c|l|r|}
     \hline
        & \textbf{Sentence} & $p(g_i|S)$ \\
     \hline % Los pescados grandes peque\~{n}o come
      \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
                         $\tau(g_{1},S_1)$ & El pez grande se come el peque\~{n}o &  0.831 \\
                         $\tau(g_{2},S_1)$ & El pescado grande se come el peque\~{n}o & 0.169 \\
     \hline % El padre el pescado nos ha preparado para cenar
      \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
                         $\tau(g_{1},S_2)$ & Padre nos ha preparado pescado para cenar & 0.980 \\
                         $\tau(g_{2},S_2)$ & Padre nos ha preparado pez para cenar & 0.020 \\
     \hline % Aqu\'{\i} el pescado muy es dulce
     $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
                         $\tau(g_{1},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.980 \\
                         $\tau(g_{2},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.020 \\
     \hline % Un pescado grande del mar es
     \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
                         $\tau(g_{1},S_4)$ & Es un pez grande del mar & 0.596 \\
                         $\tau(g_{2},S_4)$ & Es un pescado grande del mar & 0.404 \\
     \hline % El pescado grande por esos serv\'{\i}an para alimentar
     $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
                         $\tau(g_{1},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.876 \\
                         $\tau(g_{2},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.124 \\
     \hline
   \end{tabular}
\end{center}
The scores are normalised to produce fractional counts.
}

\block[titleinnersep=10em]{Model}{

Extract $n$-gram contexts and train MaxEnt model using fractional counts.

\begin{center}
\begin{tabular}{|l|r|l|r|}
\hline
$h^s(t,c)$ & $\lambda$ & $h^s(t,c)$ & $\lambda$ \\
\hline
arrain:pez & 1.1356 & arrain:pescado & 0.90723 \\
aita arrain:pez & 0.167523 & aita arrain:pescado & 1.28196 \\
arrain:pez handi & 1.37525 & arrain:pescado handi & 0.49622 \\ 
arrain:pez oso & 0.167523 & arrain:pescado oso & 1.28196  \\ 
arrain:pez prestatu & 0.167523 & arrain:pescado prestatu & 1.28196  \\
hemen arrain:pez & 0.167523 & hemen arrain:pescado & 1.28196 \\ 
itsaso arrain:pez & 0.8466  & itsaso arrain:pescado & 1.40647  \\

@@

\block[titleinnersep=10em]{\emph{Gaur arraina prestatuko dut.} `Today I'll cook fish.'}{
%Compile features into finite-state transducer. 
  \begin{multicols*}{2}
\textbf{Input:} (SL $\rightarrow$ TL)
~\\
~\\
\begin{tabular}{ll}
~~  &  \texttt{gaur<adv>} $\rightarrow$ \texttt{hoy<adv>}  \\ 
~~  &  \texttt{arrain<n>} $\rightarrow$ \texttt{pescado<n><m><ND> | pez<n><m><ND>} \\ 
~~  &  \texttt{prestatu<vblex><pfut>} $\rightarrow$ \texttt{preparar<vblex><pfut>} \\
~~  &  \texttt{ukan<vbsint><pri><NR\_HU><NK\_NI>} $\rightarrow$ \texttt{tener<vblex><pri>}
\end{tabular}
~\\
~\\
~\\
Paths in the rule transducer are traversed and the weights of \\
matching features are summed:
\begin{displaymath}
p_s(t|c) = \frac{1}{Z} \exp{ \sum_{k=1}^{n_F} \lambda_k^s h_k^s(t, c)}
\label{eq:max-ent-sum}
\end{displaymath}

    arrain:pez  = 1.1356 + 0.16752 = \textbf{1.303123}

@@

\begin{center}
      \textbf{Rule transducer}
\end{center}

@@

%\block[titleinnersep=10em]{Training}{
%
%\begin{alltt}
%0 \$ 0.830 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%1 \$ 0.134 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%1 \$ 0.034 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%0 \$ 0.001 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%0 \$ 0.015 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%0 \$ 0.005 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%1 \$ 0.846 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%1 \$ 0.134 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%\end{alltt}
%\ldots
%}
%
\column{0.55}

\block[titleinnersep=10em]{Result comparison}{

@@

\block[titleinnersep=10em]{Results}{ % See Section 4.2
 \begin{center}

@@

 \end{center}
 \texttt{Ling}: Linguistic defaults; \texttt{LER} = Lexical-selection error rate.

>>>>

\settitle{ 
  \vbox{
    \color{titlefgcolor} {\bfseries \Huge  \@title \par}
%    \vspace*{-0.2em}
    \begin{tabular}{llr}
    {\Large \@author \par} & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  &  \multirow{3}{*}{\@titlegraphic } \\ % [\TP@titlegraphictotitledistance]}  \\
    & ~ & ~   \\
    {\large \@institute}   & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ &  \\
    \end{tabular}
    \vspace*{1em}
  } 
}

@@

\title{Unsupervised training of maximum-entropy models 
 for lexical selection in rule-based machine translation}
 \institute{$^\star$UiT Norgga árktalaš universitehta, $^\dagger$Universitat d'Alacant} % See Section 4.1
\author{Francis M. Tyers,$^\star$ Felipe Sánchez-Martínez,$^\dagger$ Mikel L. Forcada$^\dagger$} 
\titlegraphic{
\includegraphics[width=0.04\textwidth]{LogoSamisk.png} \hspace{35pt} \includegraphics[width=0.15\textwidth]{logodlsioficialcolor.eps} \hspace{15pt} \includegraphics[width=0.05\textwidth,angle=90]{apertium5a.pdf}
}
%\usetheme{Basic} % See Section 5
%\usebackgroundstyle{Rays}
\begin{document}

@@

\block[titleinnersep=14em]{Lexical selection}{

@@

  \block[titleinnersep=14em]{Apertium (\url{www.apertium.org})}{
\vspace{-1pt}
    \begin{center}
      \includegraphics[width=0.45\textwidth]{architecture-lexsel.pdf}
    \end{center}
    \begin{itemize}
      \item Free/open-source rule-based MT platform
      \item Around 40 released language pairs
    \end{itemize}
  }
% -- weights should go on operations
%  <rule weight="0.906306">
%    <match lemma="ondare" tags="n"><select lemma="acervo" tags="n.*"/></match>
%    <match lemma="historiko" tags="adj.*"/>
%  </rule>
%  <rule weight="1.08987">
%    <match lemma="ondare" tags="n"><select lemma="patrimonio" tags="n.*"/></match>
%    <match lemma="historiko" tags="adj.*"/>
%  </rule>
%

  \block[titleinnersep=14em]{Schema}{

%This is produced in the first stages of RBMT, just after morphological
%analysis, part-of-speech tagging, and bilingual dictionary lookup, and
%before any structural transfer takes place (we will call this
%\emph{pre-lexsel}). In our model, it is after these first stages that
%lexical selection (\emph{lexsel}) occurs. After lexical selection,
%structural transfer and generation take place; a function
%$\tau(g_i,S)$ represents the result of these last stages, which we
%will call \emph{post-lexsel}, and returns a finished translation of a
%specific lexical-selection path $g_i$ of sentence
%

\begin{displaymath}
    S \to \mbox{\framebox{\parbox{4.0cm}{\textit{pre-lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{3.5cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{4.0cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
  \end{displaymath}
\vspace*{0.1pt}
\begin{itemize}
 \item \textbf{pre-lexsel:} pipeline up to lexical transfer
 \item \textbf{lexsel:} selects one of them \(g^\star\)
 \item \textbf{post-lexsel:} returns finished translation \(\tau(g^\star,S)\)
\end{itemize}
%Source sentence \(S\) has \(|G|\) lexical selection paths \(g_i\): \emph{lexsel} selects one of them \(g^\star\), which is used to generate translation \(\tau(g^\star,S)\).

%  \begin{center}
%    \includegraphics[width=0.40\textwidth]{schema.eps}
%  \end{center}
  }

@@

\block[titleinnersep=14em]{Training}{ % See Section 4.2

Translate SL sentences to TL using existing MT system and score the translations using an $n$-gram TL model.
~\\
\useinnerblockstyle{Table}
\begin{center}
\begin{tabular}{|c|l|r|}
     \hline
        & \textbf{Sentence} & $p(g_i|S)$ \\
     \hline % Los pescados grandes peque\~{n}o come
      \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
                         $\tau(g_{1},S_1)$ & El pez grande se come el peque\~{n}o &  0.831 \\
                         $\tau(g_{2},S_1)$ & El pescado grande se come el peque\~{n}o & 0.169 \\
     \hline % El padre el pescado nos ha preparado para cenar
      \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
                         $\tau(g_{1},S_2)$ & Padre nos ha preparado pescado para cenar & 0.980 \\
                         $\tau(g_{2},S_2)$ & Padre nos ha preparado pez para cenar & 0.020 \\
     \hline % Aqu\'{\i} el pescado muy es dulce
     $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
                         $\tau(g_{1},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.980 \\
                         $\tau(g_{2},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.020 \\
     \hline % Un pescado grande del mar es
     \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
                         $\tau(g_{1},S_4)$ & Es un pez grande del mar & 0.596 \\
                         $\tau(g_{2},S_4)$ & Es un pescado grande del mar & 0.404 \\
     \hline % El pescado grande por esos serv\'{\i}an para alimentar
     $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
                         $\tau(g_{1},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.876 \\
                         $\tau(g_{2},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.124 \\
     \hline
   \end{tabular}
\end{center}
The scores are normalised to produce fractional counts.
}

\block[titleinnersep=14em]{Model}{

Extract $n$-gram contexts and train MaxEnt model using fractional counts.
\vspace{0.2pt}
\begin{center}
\begin{tabular}{|l|r|l|r|}
\hline
$h^s(t,c)$ & $\lambda$ & $h^s(t,c)$ & $\lambda$ \\
\hline
arrain:pez & 1.1356 & arrain:pescado & 0.90723 \\
aita arrain:pez & 0.167523 & aita arrain:pescado & 1.28196 \\
arrain:pez handi & 1.37525 & arrain:pescado handi & 0.49622 \\ 
arrain:pez oso & 0.167523 & arrain:pescado oso & 1.28196  \\ 
arrain:pez prestatu & 0.167523 & arrain:pescado prestatu & 1.28196  \\
hemen arrain:pez & 0.167523 & hemen arrain:pescado & 1.28196 \\ 
itsaso arrain:pez & 0.8466  & itsaso arrain:pescado & 1.40647  \\

@@

\block[titleinnersep=14em]{\emph{Gaur arraina prestatuko dut.} `Today I'll cook fish.'}{
%Compile features into finite-state transducer. 
  \begin{multicols*}{2}
\textbf{Input:} (SL $\rightarrow$ TL)
~\\
~\\
\begin{tabular}{ll}
~~  &  \texttt{gaur<adv>} $\rightarrow$ \texttt{hoy<adv>}  \\ 
~~  &  \texttt{arrain<n>} $\rightarrow$ \texttt{pescado<n><m><ND> | pez<n><m><ND>} \\ 
~~  &  \texttt{prestatu<vblex><pfut>} $\rightarrow$ \texttt{preparar<vblex><pfut>} \\
~~  &  \texttt{ukan<vbsint><pri><NR\_HU><NK\_NI>} $\rightarrow$ \texttt{tener<vblex><pri>}
\end{tabular}
~\\
~\\
~\\
Paths in the rule transducer are traversed and the weights of \\
matching features are summed:
\begin{displaymath}
p_s(t|c) = \frac{1}{Z} \exp{ \sum_{k=1}^{n_F} \lambda_k^s h_k^s(t, c)}
\label{eq:max-ent-sum}
\end{displaymath}

    arrain:pez  = 1.1356 + 0.16752 = 1.303123

@@

\begin{center}
      \textbf{Rule transducer} \\
{\small {\tt <:$n$>} denotes rule number}
\end{center}

@@

%\block[titleinnersep=14em]{Training}{
%
%\begin{alltt}
%0 \$ 0.830 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%1 \$ 0.134 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%1 \$ 0.034 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%0 \$ 0.001 \# arrain:0 arrain\_handi:0 \# arrain:1 arrain\_handi:1  \# \\
%0 \$ 0.015 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%0 \$ 0.005 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%1 \$ 0.846 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%1 \$ 0.134 \# arrain:0 aita\_arrain:0 arrain\_prestatu:0 \# arrain:1 aita\_arrain:1 arrain\_prestatu:1 \# \\
%\end{alltt}
%\ldots
%}
%
\column{0.55}

\block[titleinnersep=14em]{Result comparison}{

@@

\block[titleinnersep=14em]{Results}{ % See Section 4.2
 \begin{center}

@@

 \end{center}
~\\
 \texttt{Ling}: Linguistic defaults; \texttt{LER} = Lexical-selection error rate.
